---
title: "data visualisation"
author: "wyne maina"
date: "2026-03-13"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## LOAD THE NECESSARY PACKAGE
```{r}
library("dplyr")
library("lubridate")
library("ggplot2")
library("tidyr")
```
## SET THE WORKING DIRECTORY
```{r}
setwd("C:/Users/Admin/Desktop/DML")
```
## LOAD THE WORKING DIRECTORY
```{r}
library(readr)
df<- read_csv("StudentsPerformance.csv")
View(df)
```
## LOAD THE NECESSARY PACKAGE
```{r}
# First, load the dataset
students <- read.csv("StudentsPerformance.csv")

# Fix column names (remove spaces)
names(students) <- make.names(names(students))

# Look at the data
head(students)
```
```{r}
colnames(data)
```
##Question 1: Line Graph - Average Scores by Parental Education
```{r}

# Q1: Calculate averages
q1_data <- summarise(group_by(students, parental.level.of.education),
                     Math = mean(math.score),
                     Reading = mean(reading.score),
                     Writing = mean(writing.score))

# Convert to long format
q1_long <- pivot_longer(q1_data, 
                        cols = c("Math", "Reading", "Writing"),
                        names_to = "Subject",
                        values_to = "Average_Score")

# Create plot
ggplot(q1_long, aes(x = parental.level.of.education, 
                    y = Average_Score, 
                    group = Subject,
                    color = Subject)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Average Scores by Parental Education",
       x = "Parental Education Level",
       y = "Average Score")
```
# Q2: Calculate averages by test prep
```{r}
# Calculate averages by test prep
q2_data <- students %>%
  group_by(test.preparation.course) %>%
  summarise(
    Math = mean(math.score),
    Reading = mean(reading.score),
    Writing = mean(writing.score)
  ) %>%
  pivot_longer(cols = c(Math, Reading, Writing),
               names_to = "Subject",
               values_to = "Average_Score")

# Create line graph
ggplot(q2_data, aes(x = Subject, 
                    y = Average_Score, 
                    group = test.preparation.course,
                    color = test.preparation.course)) +
  geom_line(size = 1) +
  geom_point(size = 4) +
  theme_minimal() +
  labs(title = "Test Preparation Course Comparison",
       x = "Subject",
       y = "Average Score",
       color = "Test Prep")

# Check if preparation helps consistently
q2_comparison <- q2_data %>%
  pivot_wider(names_from = test.preparation.course, 
              values_from = Average_Score) %>%
  mutate(Improvement = completed - none)

print("Score improvement from test prep:")
print(q2_comparison)
```
# Question 3: Grouped Bar Chart - Math Score by Gender and Test Prep
```{r}
# Calculate average math scores
q3_data <- students %>%
  group_by(gender, test.preparation.course) %>%
  summarise(avg_math = mean(math.score), .groups = 'drop')

# Create grouped bar chart
ggplot(q3_data, aes(x = gender, 
                    y = avg_math, 
                    fill = test.preparation.course)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  geom_text(aes(label = round(avg_math, 1)), 
            position = position_dodge(0.7), 
            vjust = -0.3) +
  theme_minimal() +
  labs(title = "Average Math Score by Gender and Test Prep",
       x = "Gender",
       y = "Average Math Score",
       fill = "Test Prep")

# Find highest group
q3_highest <- q3_data %>%
  arrange(desc(avg_math)) %>%
  slice(1)

print("Group with highest math score:")
print(q3_highest)
```

```{r}
# Count students by parental education and lunch type
q4_data <- students %>%
  group_by(parental.level.of.education, lunch) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(parental.level.of.education) %>%
  mutate(proportion = count / sum(count))

# Create stacked bar chart
ggplot(q4_data, aes(x = parental.level.of.education, 
                    y = proportion, 
                    fill = lunch)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Lunch Type Distribution by Parental Education",
       x = "Parental Education",
       y = "Proportion",
       fill = "Lunch Type") +
  scale_y_continuous(labels = scales::percent)

# Find highest free/reduced lunch group
q4_highest <- q4_data %>%
  filter(lunch == "free/reduced") %>%
  arrange(desc(proportion))

print("Highest free/reduced lunch proportion:")
print(head(q4_highest, 1))
```
##Question 5: Scatter Plot - Math vs Reading Score
```{r}
# Create scatter plot
ggplot(students, aes(x = math.score, y = reading.score, 
                     color = gender)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Math vs Reading Scores by Gender",
       x = "Math Score",
       y = "Reading Score",
       color = "Gender")

# Calculate correlations by gender
q5_correlations <- students %>%
  group_by(gender) %>%
  summarise(
    correlation = cor(math.score, reading.score)
  )

print("Correlation by gender:")
print(q5_correlations)
```
##Question 6: Scatter Plot - Writing vs Reading Score (Faceted)
```{r}
# Create faceted scatter plot
ggplot(students, aes(x = writing.score, y = reading.score)) +
  geom_point(alpha = 0.3, color = "darkblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  facet_wrap(~test.preparation.course) +
  theme_minimal() +
  labs(title = "Writing vs Reading Scores by Test Prep",
       x = "Writing Score",
       y = "Reading Score")

# Calculate correlations by test prep
q6_correlations <- students %>%
  group_by(test.preparation.course) %>%
  summarise(
    correlation = cor(writing.score, reading.score)
  )

print("Correlation by test prep:")
print(q6_correlations)
```
##Question 7: Histogram - Math Score Distribution by Gender
```{r}
# Create faceted histogram
ggplot(students, aes(x = math.score)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  facet_wrap(~gender) +
  theme_minimal() +
  labs(title = "Math Score Distribution by Gender",
       x = "Math Score",
       y = "Count") +
  geom_vline(data = students %>% 
               group_by(gender) %>% 
               summarise(mean_math = mean(math.score)),
             aes(xintercept = mean_math), 
             color = "red", linetype = "dashed")

# Summary statistics
q7_stats <- students %>%
  group_by(gender) %>%
  summarise(
    Mean = mean(math.score),
    Median = median(math.score),
    SD = sd(math.score),
    Min = min(math.score),
    Max = max(math.score)
  )

print("Math score statistics by gender:")
print(q7_stats)
```
##Question 8: Histogram - Writing Scores by Lunch Type
```{r}
# Create overlapping histograms
ggplot(students, aes(x = writing.score, fill = lunch)) +
  geom_histogram(bins = 20, position = "identity", alpha = 0.5, color = "white") +
  theme_minimal() +
  labs(title = "Writing Score Distribution by Lunch Type",
       x = "Writing Score",
       y = "Count",
       fill = "Lunch Type") +
  geom_vline(data = students %>% 
               group_by(lunch) %>% 
               summarise(mean_write = mean(writing.score)),
             aes(xintercept = mean_write, color = lunch), 
             linetype = "dashed", size = 1,
             show.legend = FALSE)

# Summary statistics
q8_stats <- students %>%
  group_by(lunch) %>%
  summarise(
    Mean = mean(writing.score),
    Median = median(writing.score),
    SD = sd(writing.score)
  )

print("Writing score statistics by lunch type:")
print(q8_stats)
```
##Question 9: Faceted Histogram - Reading Scores by Parental Education
```{r}
# Create faceted histogram
ggplot(students, aes(x = reading.score)) +
  geom_histogram(bins = 20, fill = "coral", color = "white") +
  facet_wrap(~parental.level.of.education, ncol = 3) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 8)) +
  labs(title = "Reading Score Distribution by Parental Education",
       x = "Reading Score",
       y = "Count")

# Calculate mean and median to assess skewness
q9_skew <- students %>%
  group_by(parental.level.of.education) %>%
  summarise(
    Mean = mean(reading.score),
    Median = median(reading.score),
    Count = n()
  ) %>%
  mutate(
    Skewness = ifelse(Mean > Median, "Right-skewed", 
                     ifelse(Mean < Median, "Left-skewed", "Symmetric"))
  ) %>%
  arrange(desc(Mean - Median))

print("Skewness assessment:")
print(q9_skew)
```
##Question 10: Histogram - Average Student Score
```{r}
# Create average score variable
students <- students %>%
  mutate(average_score = (math.score + reading.score + writing.score) / 3)

# Calculate mean
mean_avg <- mean(students$average_score)

# Create histogram
ggplot(students, aes(x = average_score)) +
  geom_histogram(bins = 20, fill = "purple", color = "white") +
  geom_vline(xintercept = mean_avg, 
             color = "red", size = 1.5, linetype = "dashed") +
  theme_minimal() +
  labs(title = "Distribution of Average Student Scores",
       subtitle = paste("Mean =", round(mean_avg, 2),
                        "| Median =", round(median(students$average_score), 2)),
       x = "Average Score (Math + Reading + Writing)/3",
       y = "Count")

# Summary statistics
q10_stats <- students %>%
  summarise(
    Mean = mean(average_score),
    Median = median(average_score),
    SD = sd(average_score),
    Min = min(average_score),
    Max = max(average_score)
  )

print("Average score statistics:")
print(q10_stats)

# QQ plot for normality check
ggplot(students, aes(sample = average_score)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal() +
  labs(title = "Q-Q Plot of Average Scores",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles")
```