---
title: "working data"
author: "wine maina"
date: "2026-03-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## IMPORT THE NECESSARY PACKAGES
```{r}
library("dplyr")
library("lubridate")
```
## SETTING THE WORKING DIRECTORY
```{r}
setwd("C:/Users/Admin/Desktop/My project")
```
## LOAD THE WORKING DIRECTORY
```{r}
library(readr)
 read_csv("lubridate_dates_project_dataset.csv")
```
##QUESTION 1.CREATE NEW COLUMN CALLED order_quarter THAT SHOWS THE QUARTER IN WHICH THE ORDER WAS PLACED
```{r}
colnames(df)
df<-df%>%
  mutate(
    #convert to date format
    order_date=ymd(Order_Date),
    #create the quarter column
    order_quarter=
      quarter(order_date)
  )
print(df)
```
#QUESTION 2.NUMBER OF ORDERS PLACED IN EACH QUARTER
```{r}
df_summary<-df%>%
  count(order_quarter)
print(df_summary)
```
##QUESTION 3.FIND THE EARLIEST ORDER DATE AN THE LATEST ORDER DATE
```{r}
df%>%
  summarise(earliest_order=min(order_date),
            latest_order=max(order_date))
print(df)
```
#QUESTION 4.FIND THE NUMBER OF ORDERS PLACED IN EACH MONTH
```{r}
df%>%
  mutate(order_month=month(order_date,label=TRUE))%>%
  count(order_month)
print(df)
```
#QUESTION 5.FIND THE AVERAGE ORDER AMOUNT FOR EACH WEEKDAY
```{r}
df%>%
  mutate(order_weekday=wday(order_date,label=TRUE))%>%
  group_by(order_weekday)%>%
  summarise(avg_order_amount=mean(order_amount))
print(df)
```
#QUESTION 6.CREATE ORDER_WEEK COLUMN (WEEK NUMBER OF YEAR)
```{r}
df<-df%>%
  mutate(order_week=week(order_date))
print(df)
```
#QUESTION 7.FIND THE NUMBER OF ORDERS PLACED IN EACH WEEK
```{r}
df%>%
  count(order_week)
print(df)
```
#QUESTION 8.FIND THE TOTAL SALES FOR EACH QUARTER 
```{r}
df%>%
  group_by(order_quarter)%>%
  summarise(total_sales=sum(order_amount))
print(df)
```