---
title: "student perfomance histogram"
author: "wyne maina"
date: "2026-03-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### LOAD THE REGUIRED PACKAGE
```{r}
# First, load the dataset
students <- read.csv("StudentsPerformance.csv")

# Check if there are any missing values (this dataset has none)
print("Checking for missing values:")
print(sum(is.na(students)))

# Since there are no missing values, let's create a version WITH some missing values for practice
# This is just for learning - you wouldn't do this with real data!
set.seed(123)
students_na <- students
students_na$math.score[sample(1:nrow(students_na), 30)] <- NA
students_na$reading.score[sample(1:nrow(students_na), 20)] <- NA
students_na$writing.score[sample(1:nrow(students_na), 25)] <- NA

# Now let's solve each question using this dataset with missing values
```
##Q1. Count missing values using is.na() and colSums()
```{r}
# Simple code for Q1
total_missing <- sum(is.na(students_na))
print(paste("Total missing values:", total_missing))

# Missing values per column
missing_per_column <- colSums(is.na(students_na))
print("Missing values per column:")
print(missing_per_column)

# Sort in descending order
print("Sorted descending:")
print(sort(missing_per_column, decreasing = TRUE))
```
##Q2. Calculate percentage of missing values
```{r}
# Simple code for Q2
total_rows <- nrow(students_na)
missing_pct <- (colSums(is.na(students_na)) / total_rows) * 100
print("Missing percentage per column:")
print(round(missing_pct, 2))

# Show only columns with >10% missing
print("Columns with >10% missing:")
print(missing_pct[missing_pct > 10])
```
##Q3. Count rows with and without missing values
```{r}
# Simple code for Q3
rows_with_na <- sum(!complete.cases(students_na))
rows_complete <- sum(complete.cases(students_na))

print(paste("Rows with at least one missing value:", rows_with_na))
print(paste("Rows with no missing values:", rows_complete))
```
##Q4. Create custom summary table
```{r}
# Simple code for Q4
summary_table <- data.frame(
  Column = names(students_na),
  Data_Type = sapply(students_na, class),
  Total = nrow(students_na),
  Non_Missing = sapply(students_na, function(x) sum(!is.na(x))),
  Missing = sapply(students_na, function(x) sum(is.na(x)))
)

print("Custom Summary Table:")
print(summary_table)
```
##
Q5. Create barplot of missing values
```{r}
# Simple code for Q5
missing_counts <- colSums(is.na(students_na))

# Only show columns with missing values
missing_counts <- missing_counts[missing_counts > 0]

# Create barplot
barplot(missing_counts,
        main = "Missing Values per Column",
        ylab = "Number Missing",
        col = "skyblue",
        las = 2)  # Makes labels vertical
```
##Q6. Filter rows where math.score and reading.score are not missing
```{r}
# Simple code for Q6
complete_scores <- students_na[!is.na(students_na$math.score) & 
                                !is.na(students_na$reading.score), ]

print(paste("Rows with complete math and reading scores:", nrow(complete_scores)))
```
##Q7. Find rows with missing values in any score column
```{r}
# Simple code for Q7
missing_in_scores <- students_na[is.na(students_na$math.score) | 
                                  is.na(students_na$reading.score) | 
                                  is.na(students_na$writing.score), ]

print("First 10 rows with missing scores (showing row numbers):")
print(head(missing_in_scores[, c("gender", "math.score", "reading.score", "writing.score")], 10))
print(paste("Total such rows:", nrow(missing_in_scores)))
```
##Q8. Filter female students with complete math scores
```{r}
# Simple code for Q8
female_complete <- students_na[students_na$gender == "female" & 
                                !is.na(students_na$math.score), ]

print(paste("Female students with complete math scores:", nrow(female_complete)))
```
##Q9. Replace missing values with mean and "Unknown"
```{r}
# Simple code for Q9
# Create a copy
students_fixed <- students_na

# Replace missing math scores with mean
mean_math <- mean(students_na$math.score, na.rm = TRUE)
students_fixed$math.score[is.na(students_fixed$math.score)] <- round(mean_math, 1)

# Check if any NAs remain in math.score
print(paste("NAs in math.score after fix:", sum(is.na(students_fixed$math.score))))
```
##Q10. Replace missing values using simple methods
```{r}
# Simple code for Q10
# Create final cleaned dataset
students_clean <- students_na

# Calculate values for replacement
median_math <- median(students_na$math.score, na.rm = TRUE)
median_reading <- median(students_na$reading.score, na.rm = TRUE)
mean_writing <- mean(students_na$writing.score, na.rm = TRUE)

# Replace missing values
students_clean$math.score[is.na(students_clean$math.score)] <- median_math
students_clean$reading.score[is.na(students_clean$reading.score)] <- median_reading
students_clean$writing.score[is.na(students_clean$writing.score)] <- round(mean_writing, 1)

# Verify no NAs remain
print("Checking for remaining NAs:")
print(paste("Math NAs:", sum(is.na(students_clean$math.score))))
print(paste("Reading NAs:", sum(is.na(students_clean$reading.score))))
print(paste("Writing NAs:", sum(is.na(students_clean$writing.score))))
print(paste("Total NAs in cleaned dataset:", sum(is.na(students_clean))))
```