---
title: "STUDENT ANALYSIS"
author: "wyne maina"
date: "2026-03-18"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## LOAD THE NECESSARY PACKAGE
```{r}
library("dplyr")
library("lubridate")
library("ggplot2")
library("tidyverse")
```

```{r}
library(tidyverse)

# Load the dataset
df <- read.csv("StudentsPerformance.csv")

# Clean column names for easier coding (replaces spaces and slashes with underscores)
colnames(df) <- gsub("\\.", "_", make.names(colnames(df)))
# Resulting names: gender, race_ethnicity, parental_level_of_education, lunch, 
# test_preparation_course, math_score, reading_score, writing_score
```
##QUESTION 1.MATH SCORES BY PARENTAL EDUCATION
```{r}
ggplot(df, aes(x = parental_level_of_education, y = math_score)) +
  geom_boxplot(fill = "skyblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Math Scores by Parental Level of Education",
       x = "Parental Education", y = "Math Score")

# Identify median scores
df %>%
  group_by(parental_level_of_education) %>%
  summarise(median_math = median(math_score)) %>%
  arrange(desc(median_math))
```
##QUESTION 2.WRITING SCORES BY GENDER AND LUNCH TYPE
```{r}
ggplot(df, aes(x = gender, y = writing_score, fill = lunch)) +
  geom_boxplot() +
  labs(title = "Writing Scores by Gender and Lunch Type",
       x = "Gender", y = "Writing Score")
```
##QUESTION 3.READING SCORES (COMPLETED PREP)BY RAC/ETHNICITY
```{r}
prep_completed <- df %>% filter(test_preparation_course == "completed")

ggplot(prep_completed, aes(x = race_ethnicity, y = reading_score)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Reading Scores by Race (Test Prep Completed)")

# Calculate IQR to identify largest variability
prep_completed %>%
  group_by(race_ethnicity) %>%
  summarise(IQR_reading = IQR(reading_score)) %>%
  arrange(desc(IQR_reading))
```
##QUESTION 4.FACETED MATHS SCORES BY PREPARATION STATUS
```{r}
ggplot(df, aes(x = gender, y = math_score)) +
  geom_boxplot(fill = "gold") +
  facet_wrap(~test_preparation_course) +
  labs(title = "Math Score Distribution by Gender and Prep Course")
```
##QUESTION 5.SUBJECT COMPARISON
```{r}
df_long <- df %>%
  pivot_longer(cols = ends_with("score"), names_to = "Subject", values_to = "Score")

ggplot(df_long, aes(x = Subject, y = Score, fill = Subject)) +
  geom_boxplot() +
  labs(title = "Score Distributions Across Subjects")

# Check standard deviation for variability
df_long %>%
  group_by(Subject) %>%
  summarise(SD_Score = sd(Score))
```
##QUESTION 6.REDING SCORES ORDERED BY MEDIAN
```{r}
# Order factor levels based on median reading score
df_ordered <- df %>%
  mutate(parental_level_of_education = fct_reorder(parental_level_of_education, reading_score, .fun = median))

ggplot(df_ordered, aes(x = parental_level_of_education, y = reading_score)) +
  geom_boxplot(fill = "coral") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Reading Scores Ordered by Median")
```
##QUESTION 7.POOR MATHS PERFOMANCE(<-40) AND WRITING
```{r}
low_math <- df %>% filter(math_score < 40)

ggplot(low_math, aes(x = gender, y = writing_score)) +
  geom_boxplot(fill = "plum") +
  labs(title = "Writing Scores of Students with Math Score < 40")
```
##QUESTION 8.MATH OUTLIERS BY RACE/ETHNICITY
```{r}
ggplot(df, aes(x = race_ethnicity, y = math_score)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 16, outlier.size = 2) +
  labs(title = "Math Scores and Outliers by Race/Ethnicity")
```
##QUESTION 9.FACETED WRITING SCORES BY LUNCH TYPE
```{r}
ggplot(df, aes(x = gender, y = writing_score)) +
  geom_boxplot(fill = "lightblue") +
  facet_grid(. ~ lunch) +
  labs(title = "Writing Scores Faceted by Lunch Type")
```
##QUESTION 10.MATHS SCORES FOR HIGH PERFOMING READERS
```{r}
med_reading <- median(df$reading_score)
high_readers <- df %>% filter(reading_score > med_reading)

ggplot(high_readers, aes(x = test_preparation_course, y = math_score)) +
  geom_boxplot(fill = "lightgrey") +
  labs(title = "Math Scores of High-Performing Readers by Prep Status")
```