---
title: "temperature line graph"
author: "wyne maina"
date: "2026-03-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## LOAD THE REQUIRED PACKAGE
```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
```

## LOAD AND PREPARE THE DATASET
```{r}
data <- read.csv("temperature.csv")

# Convert Year column to date
data$Date <- ymd(paste0(data$Year, "-01"))

# Extract year and month
data <- data %>%
  mutate(
    Year = year(Date),
    Month = month(Date)
  )
```
##3. Global Mean Temperature Trend Over the Years
```{r}
ggplot(data, aes(x = Date, y = Mean)) +
  geom_line() +
  labs(title = "Global Mean Temperature Trend Over Time",
       x = "Year",
       y = "Temperature Anomaly") +
  theme_minimal()
```
##4. Temperature Trend for Each Month Across Years
```{r}
ggplot(data, aes(x = Year, y = Mean, color = factor(Month))) +
  geom_line() +
  labs(title = "Temperature Trend for Each Month Across Years",
       x = "Year",
       y = "Temperature Anomaly",
       color = "Month") +
  theme_minimal()
```
##5. Average Temperature Per Year
```{r}
year_avg <- data %>%
  group_by(Year) %>%
  summarise(Average_Temp = mean(Mean))

ggplot(year_avg, aes(x = Year, y = Average_Temp)) +
  geom_line(color = "blue") +
  labs(title = "Average Temperature per Year",
       x = "Year",
       y = "Average Temperature") +
  theme_minimal()
```
##6. January Temperature Trend Over the Years
```{r}
jan_data <- data %>%
  filter(Month == 1)

ggplot(jan_data, aes(x = Year, y = Mean)) +
  geom_line(color = "red") +
  labs(title = "January Temperature Trend Over Years",
       x = "Year",
       y = "Temperature Anomaly") +
  theme_minimal()
```
##6. January Temperature Trend Over the Years
```{r}
jan_data <- data %>%
  filter(Month == 1)

ggplot(jan_data, aes(x = Year, y = Mean)) +
  geom_line(color = "red") +
  labs(title = "January Temperature Trend Over Years",
       x = "Year",
       y = "Temperature Anomaly") +
  theme_minimal()
```
##7. Temperature Changes Before and After 1980
```{r}
data <- data %>%
  mutate(Period = ifelse(Year < 1980, "Before 1980", "After 1980"))

period_avg <- data %>%
  group_by(Year, Period) %>%
  summarise(Avg_Temp = mean(Mean))

ggplot(period_avg, aes(x = Year, y = Avg_Temp, color = Period)) +
  geom_line() +
  labs(title = "Temperature Changes Before and After 1980",
       x = "Year",
       y = "Average Temperature") +
  theme_minimal()
```
##8. July Temperature Trend Over the Years
```{r}
july_data <- data %>%
  filter(Month == 7)

ggplot(july_data, aes(x = Year, y = Mean)) +
  geom_line(color = "orange") +
  labs(title = "July Temperature Trend Over Years",
       x = "Year",
       y = "Temperature Anomaly") +
  theme_minimal()
```
##9. Temperature Changes from January to December for a Selected Year
```{r}
selected_year <- max(data$Year)

year_data <- data %>%
  filter(Year == selected_year)

ggplot(year_data, aes(x = Month, y = Mean)) +
  geom_line() +
  labs(title = paste("Temperature Change from January to December in", selected_year),
       x = "Month",
       y = "Temperature Anomaly") +
  theme_minimal()
```
##10. Compare Temperature Trends of Two Months (January vs July)
```{r}
two_months <- data %>%
  filter(Month %in% c(1,7))

ggplot(two_months, aes(x = Year, y = Mean, color = factor(Month))) +
  geom_line() +
  labs(title = "Comparison of January and July Temperature Trends",
       x = "Year",
       y = "Temperature Anomaly",
       color = "Month") +
  theme_minimal()
```