---
title: "WYNE"
author: "wyne maina"
date: "2026-03-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## LOAD THE NECESSARY PACKAGES
```{r}
library("dplyr")
```
## SET THE WORKING DIRECTORY
```{r}
setwd("C:/Users/Admin/Desktop/DML")
```
## LOAD THE WORKING DIRECTORY
```{r}
df<-read.csv("dplyr_conditional_project_dataset.csv")
View(df)
```

```{r}
colnames(df)
```
```{r}
library(dplyr)

# Example: Total Score = 40% Assignment + 60% Exam
df <- df %>%
  mutate(Total_Score = 0.4 * Assignment_Score + 0.6 * Exam_Score)
```
# CREATE THE GRADE COLLUMN
```{r}
df<-df%>%
  mutate(Grade=case_when(
    Total_Score>=80~"A",
    Total_Score>=70~"B",
    Total_Score>=60~"C",
    TRUE           ~"F"#this handles everything below 60
  ))
  
print(df)
```
##1 CREATE NEW COLUMN honours_category
```{r}
# Ensure you use ~ and check that Total_Score matches your csv column name exactly
df<-df%>%
  mutate(
    Total_Score=Assignment_Score+Exam_Score, # Create the missing column
    Honours_Category=case_when(
    Total_Score>=75~"First Class",
    Total_Score>=65~"Second Class Upper",
    Total_Score>=55~"Second Class Lower ",
    Total_Score>=50~"Pass",
    TRUE           ~"Fail"
    ))
    print(df)
```
#2. IDENTIFY STUDENTS ELIGIBLE FOR RETAKE
```{r}
df<-df%>%
  mutate(Retake_Required=if_else(Total_Score<50&Attendance_Percentage>70,"Yes","No"))
print(df)
```

#3.IDENTIFY YOUNG HIGH PERFOMERS
```{r}
# THIS CREATES A NEW DATAFRAME CONTAINING ONLY THE STUDENT WHO MEET BOTH CRITERIA
young_high_perfomers<-df %>%
  filter(Age<=22, Total_Score>=70)
print(young_high_perfomers)
```
#4.ATTENDANCE WARNING CATEGORY
```{r}
df <- df %>%
  mutate(Attendance_Status = case_when(
    Attendance_Percentage < 60 ~ "Critical",
    Attendance_Percentage >= 60 & Attendance_Percentage <= 74 ~ "Warning",
    Attendance_Percentage >= 75 ~ "Good Standing",
    TRUE ~ "Unknown"
  ))
print(df)

```
#5 EXTRA CREDIT ELIGIBILITY
```{r}
df<-df%>%
  mutate(Extra_Credit=if_else(Assignment_Score>85&Exam_Score>80,"Eligible","Ineligible"))
print(df)
```
#6 .DEPARTMENTALHIGH ACHIEVERS COUNT
```{r}
#Filter for grade A ,group by department, and count
high_achievers_count<-df%>%
  filter(Grade=="A")%>%
  group_by(Department)%>%
  summarise(High_Achievers_Count=n())
print(high_achievers_count)
```
#7. ACADEMIC RISK FLAG
```{r}
# %in% is used to check if a value exist within a specific vector or list of value
df<-df%>%
  mutate(Academic_Risk=case_when(
    Grade %in% c("D","F")~"At risk",
    Grade=="C"           ~"Stable",
    Grade %in% c("A","B")~"Excellent",
    TRUE                ~"Other"
  ))
print(df)
```
  ))
#8  STUDENT WHO IMPROVED IN EXAM
```{r}
# filter for rows where Exam Score is higher than Assignment Score
improved_student<-df%>%
  filter(Exam_Score>Assignment_Score)
print(improved_student)
```
#9 CONDITIONAL BONUS MARKS
```{r}
# use if_else to add 5 marks only if attendance is 90 or above
conditional_bonus<-df%>%
  mutate(Adjusted_Total=if_else( Attendance_Percentage>=90,
                                Total_Score+5,
                                Total_Score))
print(conditional_bonus)
```
# 10 OVERALL PERFOMANCE SUMMARY BY GENDER
```{r}
# group by gender and calculate pass/fail total
perfomance_summary<-df%>%
  group_by(Gender)%>%
  summarise(
    Number_Passed=sum(Total_Score>=50,na.rm = TRUE),
    Number_Failed=sum(Total_Score<50,na.rm = TRUE)
  )
print(perfomance_summary)
```
  