---
title: "data visualisation"
author: "wyne maina"
date: "2026-03-13"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## LOAD THE NECESSARY PACKAGES
```{r}
library("dplyr")
library("lubridate")
library("tidyr")
library("ggplot2")
```
## SET THE WORKING DIRECTORY
```{r}
setwd("C:/Users/Admin/Desktop/DML")
```
## LOAD THE WORKING DIRECTORY
```{r}
df<-read.csv("StudentsPerformance.csv")
```
# Load required libraries
```{r}
library(tidyverse)
library(ggplot2)
```
# Question 1: Line graph - Average scores by parental education level
```{r}
q1_data <- df %>%
  pivot_longer(cols = c(math.score, reading.score, writing.score), 
               names_to = "subject", 
               values_to = "score") %>%
  group_by(parental.level.of.education, subject) %>%
  summarise(avg_score = mean(score), .groups = 'drop') %>%
  mutate(subject = case_when(
    subject == "math.score" ~ "Math",
    subject == "reading.score" ~ "Reading", 
    subject == "writing.score" ~ "Writing"
  ))

# Reorder education levels
edu_order <- c("some high school", "high school", "some college", 
               "associate's degree", "bachelor's degree", "master's degree")
q1_data$parental.level.of.education <- factor(q1_data$parental.level.of.education, 
                                               levels = edu_order)

# Plot Q1
ggplot(q1_data, aes(x = parental.level.of.education, y = avg_score, 
                    group = subject, color = subject)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Average Scores by Parental Education Level",
       x = "Parental Education Level", 
       y = "Average Score",
       color = "Subject")

# Analysis: Which subject shows strongest improvement?
q1_improvement <- q1_data %>%
  group_by(subject) %>%
  summarise(improvement = max(avg_score) - min(avg_score)) %>%
  arrange(desc(improvement))
# Writing shows the strongest improvement
```
# Question 2: Line graph - Test preparation comparison
```{r}
q2_data <- df %>%
  pivot_longer(cols = c(math.score, reading.score, writing.score),
               names_to = "subject",
               values_to = "score") %>%
  group_by(test.preparation.course, subject) %>%
  summarise(avg_score = mean(score), .groups = 'drop') %>%
  mutate(subject = case_when(
    subject == "math.score" ~ "Math",
    subject == "reading.score" ~ "Reading",
    subject == "writing.score" ~ "Writing"
  ))

ggplot(q2_data, aes(x = subject, y = avg_score, 
                    group = test.preparation.course, 
                    color = test.preparation.course)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Average Scores by Test Preparation Course",
       x = "Subject", y = "Average Score",
       color = "Test Prep")
```
# Question 3: Grouped bar chart - Math score by gender and test prep
```{r}
q3_data <- df %>%
  group_by(gender, test.preparation.course) %>%
  summarise(avg_math = mean(math.score), .groups = 'drop')

ggplot(q3_data, aes(x = gender, y = avg_math, 
                    fill = test.preparation.course)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(title = "Average Math Score by Gender and Test Preparation",
       x = "Gender", y = "Average Math Score",
       fill = "Test Prep")

# Highest average math score group
q3_data %>% arrange(desc(avg_math))  # Female with completed prep
```
# Question 4: Stacked bar chart - Students distribution by parental education and lunch
```{r}
q4_data <- df %>%
  group_by(parental.level.of.education, lunch) %>%
  summarise(count = n(), .groups = 'drop')

q4_data$parental.level.of.education <- factor(q4_data$parental.level.of.education, 
                                               levels = edu_order)

ggplot(q4_data, aes(x = parental.level.of.education, y = count, 
                    fill = lunch)) +
  geom_col(position = "fill") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Proportion of Lunch Types by Parental Education",
       x = "Parental Education Level", y = "Proportion",
       fill = "Lunch Type")
```
# Question 5: Scatter plot - Math vs Reading with gender color
```{r}
ggplot(df, aes(x = reading.score, y = math.score, 
               color = gender, size = writing.score)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, aes(group = gender)) +
  theme_minimal() +
  labs(title = "Math vs Reading Scores by Gender",
       x = "Reading Score", y = "Math Score",
       size = "Writing Score")
```
# Question 6: Scatter plot - Writing vs Reading faceted by test prep
```{r}
ggplot(df, aes(x = reading.score, y = writing.score)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~test.preparation.course) +
  theme_minimal() +
  labs(title = "Writing vs Reading Scores by Test Preparation",
       x = "Reading Score", y = "Writing Score")
```
# Question 7: Histogram - Math score distribution by gender
```{r}
ggplot(df, aes(x = math.score)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "black") +
  facet_wrap(~gender) +
  theme_minimal() +
  labs(title = "Distribution of Math Scores by Gender",
       x = "Math Score", y = "Count")
```
# Question 8: Histogram - Writing scores by lunch type
```{r}
ggplot(df, aes(x = writing.score, fill = lunch)) +
  geom_histogram(bins = 20, position = "identity", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Distribution of Writing Scores by Lunch Type",
       x = "Writing Score", y = "Count",
       fill = "Lunch Type")
```
# Question 9: Faceted histogram - Reading scores by parental education
```{r}
ggplot(df, aes(x = reading.score)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "black") +
  facet_wrap(~parental.level.of.education) +
  theme_minimal() +
  theme(strip.text = element_text(size = 8)) +
  labs(title = "Distribution of Reading Scores by Parental Education",
       x = "Reading Score", y = "Count")
```
# Question 10: Average score distribution
```{r}
q10_data <- df %>%
  mutate(avg_score = (math.score + reading.score + writing.score) / 3)

mean_score <- mean(q10_data$avg_score)

ggplot(q10_data, aes(x = avg_score)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "black") +
  geom_vline(xintercept = mean_score, color = "red", size = 1) +
  theme_minimal() +
  labs(title = "Distribution of Average Student Scores",
       x = "Average Score", y = "Count",
       caption = "Red line indicates mean score")
```